---
title: "Reproducible Research Assignment"
author: "1066059"
date: "2023-12-03"
output:
  html_document : default
  pdf_document  : default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Installing / loading packages
```{r, echo=TRUE, results='hide'}
install.packages(c("palmerpenguins","ggplot2", "janitor", "dplyr","ragg", "svglite"))
```

```{r, message=FALSE, warning=FALSE}
library(ragg)
library(svglite)
library(tinytex)
library(palmerpenguins)
library(ggplot2)
library(janitor)
library(dplyr)
```


## QUESTION 01: Data Visualisation for Science Communication

### a) Bad Figure Example:

```{r bad figure code, echo = FALSE, warning=FALSE}
# Misleading scatter plot

plot_bad_figure <- ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm, color = species, shape = island)) +
  geom_point(size = 10, alpha = 0.7) +
  labs(title = "Penguin Body Mass vs. Flipper Length",
       x = "body_mass_g",
       y = "flipper_length_mm",
       color = "Species",
       shape = "Island") +
  theme_dark()+
  theme(legend.position = "bottom", legend.title = element_blank(),
        panel.grid = element_blank(), axis.line = element_line(color = "black"),
        plot.title = element_text(hjust = 0.5, size = 20, face = "bold", color = "red"),
        axis.title = element_text(size = 12, face = "italic", color = "purple"),
        legend.text = element_text(size = 8, face = "bold", color = "blue"))

print(plot_bad_figure)

```


### b) Write about how your design choices mislead the reader about the underlying data (200-300 words).

The aims of data visualisation involves four fundamental parts: showing the data; making patterns easy to see; representing magnitudes honestly and drawing graphical elements clearly (Whitlock and Schluter, 2020). The graph above is an example of a bad figure for many reasons. Firstly, the large data point size obscures some of the data. Secondly, the large data point size and colour choice makes it hard to determine the patterns seen in the data.The red and green colour choice is especially bad for those who are colour blind. Additionally, the shape size may influence the reader to misinterpret the magnitude of the data. To elaborate, the size and shape of the data points make it hard to determine where it correlates to on the axis and coupled with the large size of the data points, the reader may infer that the data points have a greater value than they actually do. Finally, the graphical elements are not clear and are distracting to the reader. This includes the numerous colours which are distracting and the dark theme for the background makes it harder to read as do the italicised axis titles. Also, the lack of legend titles makes it harder to interpret the graph. Overall, the graph design is not appealing, hard to read, hides the data and is misleading to the reader, therefore making it difficult to draw valid conclusions.


------------------------------------------------------------------------

## QUESTION 2: Data Pipeline

------------------------------------------------------------------------

### Introduction

This data pipeline aims to load the raw data from the Palmer penguins dataset, clean the said raw data and then perform analyses involving the production of figures. This will involve producing an exploratory figure, running statistical tests and creating a final results figure.


#### Installing / loading packages 

```{r, echo=TRUE, results='hide'}
install.packages(c("ggplot2", "palmerpenguins", "janitor", "dplyr", "ragg","ggpubr"))
library(ggplot2)
library(palmerpenguins)
library(janitor)
library(dplyr)
library(ragg)
library(ggpubr)
```


#### Loading the data 
Creating a new folder called, "data" in my project allows the raw data to be stored in it's unedited form which is beneficial for reproducibility as others can see how the data has been manipulated and it can also be used for others to use in their own analyses. 
```{r}
write.csv(penguins_raw, "penguins_raw.csv")
```


#### Cleaning the data 
I created my own cleaning functions and stored them in another folder called "functions" to prevent repetitive code being written.These functions can be called after sourcing without needing to write all of the code out again.

```{r, echo=TRUE, results='hide'}

#define function to clean columns
clean <- function(penguins_data) {
  penguins_data %>%
    select(-starts_with("Delta")) %>%
    select(-Comments) %>%
    clean_names()
}

#function to shorten species names 
shorten_species <- function(penguins_data) {
  penguins_data %>%
    mutate(species = case_when(
      species == "Adelie Penguin (Pygoscelis adeliae)" ~ "Adelie",
      species == "Chinstrap penguin (Pygoscelis antarctica)" ~ "Chinstrap",
      species == "Gentoo penguin (Pygoscelis papua)" ~ "Gentoo"
    ))
}

#function to remove empty rows / columns 
remove_empty_columns_rows <- function(penguins_data) {
  penguins_data %>%
    remove_empty(c("rows", "cols"))
}

#function to remove rows with NA values 
remove_NA <- function(penguins_data) {
  penguins_data %>%
    na.omit()
}

#piping multiple newly created functions 
penguins_clean <- penguins_raw %>%
    clean() %>%
    shorten_species() %>%
    remove_empty_columns_rows() %>%
  remove_NA
  

#checking cleaning functions worked 
names(penguins_raw)
names(penguins_clean) 

#save cleaned data 
write.csv(penguins_clean, "penguins_clean.csv")
```


#### Creating an exploratory plot 
Now that the data has been cleaned, we can start to visualise the data with exploratory plots.These plots are used to gain a basic understanding of the relationship between variables and therefore do not need to look especially pretty.
```{r Data Exploration, warning=FALSE}

#exploratory plot looking at the relationship between body mass and flipper length
plot_explore <- function(penguins_clean){
  penguins_clean %>%
  ggplot(aes(x = body_mass_g, y = flipper_length_mm , color = species)) +
  geom_point() +
  labs(title = "Exploratory Plot - Penguin Body Mass vs. Flipper Length",
       x = "Body Mass (g)",
       y = "Flipper Length (mm)")
}

exploratory_plot <- plot_explore(penguins_clean)
exploratory_plot
```

```{r, include=FALSE}
#function for saving vector image 
save_explore_plot_svg <- function(penguins_clean, 
                                  filename, size, scaling){
    size_inches = size/2.54
    svglite(filename, width   = size_inches, 
                      height  = size_inches, 
                      scaling = scaling)
    explore_plot <- plot_explore(penguins_clean)
    print(exploratory_plot)
    dev.off()
}

save_explore_plot_svg(penguins_clean, 
                      "explore_plot_vector.svg", 
                      size = 15, scaling = 1)

```


### Hypothesis
The exploratory plot shows that there is a positive correlation between body mass and flipper length. That is to say that as body mass increases, so does flipper length. The plot also shows that species cluster together which may suggesting that there may be a correlation between body mass and flipper length within species.
To look at this, I have separated the species into their own plots so that the effects of species can be seen individually. 

### Statistical Methods
SUbsetting into different species and then performing a linear regression for each will help to identify the statistical significance betweenbody mass and flipper length correlation in the different penguin species. 

```{r Statistics}

#subset into species to look at the linear models for each species 
adelie <- subset(penguins_clean, species == "Adelie")
chinstrap <- subset(penguins_clean, species == "Chinstrap")
gentoo <- subset(penguins_clean, species == "Gentoo")

# fit linear regression models to the different species 
#adelie
adelie_linear <- lm(body_mass_g ~ flipper_length_mm, data = adelie)
summary(adelie_linear) 

#chinstrap
chinstrap_linear <- lm(body_mass_g ~ flipper_length_mm, data = chinstrap)
summary(chinstrap_linear) 

#gentoo
gentoo_linear <- lm(body_mass_g ~ flipper_length_mm, data = gentoo)
summary(gentoo_linear) 

```


### Results & Discussion
The results figure shows the relationship between body mass and flipper length in the three different penguin species. The plot also shows the linear regression for each of the different species along with the Pearson correlation coefficient (R) showing the strength of the correlation and the p-value. 
All three species show significant p-values for these coefficients, suggesting that the linear regression model suggests there is a correlation between body mass and flipper length in the three species.
Additionally, the coefficients vary between species, suggesting there may be species-specific effects between body mass and flipper length correlation. 
To elaborate, in Adelie penguins, the linear regression coefficient (32.689) was shown to be significant (p-value < 3.4e-09). In chinstrap penguins, the linear regression coefficient was 34.573  and also shown to be statistically significant with a p-value of 3.75e-09. Finally, the gentoo penguins showed the strongest correlation between body mass and flipper length with a linear regression coefficient of 54.165 and a p-value of < 2e-16. 

#### Results Figure
```{r Plotting Results, warning=FALSE, message=FALSE}
#creating function for plotting 
plot_explanatory <- function(penguins_clean){
  penguins_clean %>%
    ggplot(aes(x = body_mass_g, y = flipper_length_mm , color = species)) +
geom_point() +
  facet_wrap(~species)+
    geom_smooth(method = "lm", se = FALSE, color = "black")+
    stat_cor(method = "pearson", size = 2) +
  labs(title = "Penguin Body Mass vs. Flipper Length Across Species",
       x = "Body Mass (g)",
       y = "Flipper Length (mm)")+
    theme_minimal()+
    theme(axis.title.x = element_text(margin = margin(t = 10)),
          axis.title.y = element_text(margin = margin(r = 10)),
          plot.title = element_text(hjust = 0.5))
    }

#plotting explanatory figure
explanatory_plot <- plot_explanatory(penguins_clean)
explanatory_plot
```

```{r, include=FALSE}
#saving figure as a vector image so it can be scaled 
save_explanatory_plot_svg <- function(penguins_clean, 
                                      filename, size, scaling){
  size_inches = size/2.54
  svglite(filename, width   = size_inches, 
          height  = size_inches, 
          scaling = scaling)
  explanatory_plot <- plot_explanatory(penguins_clean)
  print(explanatory_plot)
  dev.off()
}

save_explanatory_plot_svg(penguins_clean,
                          "explanatory_plot_vector.svg", 
                      size = 15, scaling = 1)
```

### Conclusion

In conclusion, the results support the hypothesis that there is a positive correlation between body mass and flipper length in  the Palmer Penguins study. Additionally, after noticing clustering of species in the exploratory plot, I subsetted the data to look at whether there were species-specific effects on the correlation between body mass and flipper length and the results also support this, with Gentoo penguins having the strongest correlation of the species. 



------------------------------------------------------------------------


## QUESTION 3: Open Science

### a) GitHub

*GitHub link:*
https://github.com/1066059/reproducible_science_assignment


### b) Share your repo with a partner, download, and try to run their data pipeline.

*Partner's GitHub link:*
https://github.com/wings029/Reproducible-science-and-figures-assignment


### c) Reflect on your experience running their code. (300-500 words)

**What elements of your partner's code helped you to understand their data pipeline?**

My partner described what they were going to do and why they were doing it before each chunk of code and also used hashtags to make comments within chunks which made it very easy to understand. The text before chunks allowed the general idea to be seen and the comments in the chunk allowed each individual action to be understood. Also the use of piping kept the code logical and neat and enhanced the reproducibility as it prevents mistakes.

**Did it run? Did you need to fix anything?**

Everything ran smoothly, there was no need to fix anything.

**What suggestions would you make for improving their code to make it more understandable or reproducible, and why?**

I suggested to keep separate subfolders for functions/figures so that they can be shared and easily applied to other data analyses.This also prevents the need for repetitive code writing which may lead to mistakes and therefore hinder reproducibility. I also suggested to save the figures as vector images rather than .png. I thought this would be useful as they can be scaled for different uses and are therefore more reproducible in print. Additionally, the final results figure involves adding the statistics as annotations/labels on the graph. To make this more reproducible, I would suggest using a function to automatically add the statistics rather than adding them in manually. This wasn't an issue as the code ran well but if I had had problems with the chunks coding the statistics section then this may have been a problem.

**If you needed to alter your partner's figure using their code, do you think that would be easy or difficult, and why?**

I did not need to alter their figure but I think it would be easy to do so if needed because the steps leading to the production of the figure were clear and logical and described well. However, it may have been a little bit difficult due to the above mentioned labelling issue with the statistics rather than using a function to do this. As already mentioned, if I hadn't run the chunk containing the statistics, the labels/annotations on the graph wouldn't make any sense.


### d) Reflect on your own code based on your experience with your partner's code and their review of yours. (300-500 words)

**What improvements did they suggest, and do you agree?**

My partner made three suggestions to improve on. Firstly, to show the original raw dataset before cleaning in order to allow those who are unfamiliar with it to gain an understanding of what the raw data looks like. I think this is a good suggestion for the future as not everyone has come across this dataset before and it would be useful to orient oneself with the variables the dataset contains before doing any analysis. For this exercise, I assumed people had previous knowledge of the raw dataset and didn't feel the need to include it at the beginning for reference but I agree that it makes sense to include it.
The second piece of advice was to get R to copy and store the functions I made rather than manually doing it. This is a good idea but providing this information in the code allows others to see the thought process and perform it for themselves and therefore helps reproducibility. 
Finally, my partner suggested to keep a record of the versions of the packages used in my analyses. This is a good suggestion as the packages may be outdated or others may have different versions which may hinder them from running the code.


**What did you learn about writing code for other people?**

I learnt that is a very useful skill to have because it not only helps others to understand the code, but also myself. There have been times when I have returned to a project and cannot remember why I took the steps I did. Writing code for other people has helped me to lay out my analyses more clearly.I also learnt the importance of breaking down my code into manageable chunks with their own specific functionality. This allows other to easily understand what and why an action is being performed and also helps with error handling. 



































