---
title: "Reproducible Research Assignment"
author: "1066059"
date: "2023-12-03"
output:
  html_document : default
  pdf_document  : default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Installing / loading packages
```{r, echo=TRUE, results='hide'}
install.packages(c("palmerpenguins","ggplot2", "janitor", "dplyr","ragg", "svglite"))
```

```{r, message=FALSE, warning=FALSE}
library(ragg)
library(svglite)
library(tinytex)
library(palmerpenguins)
library(ggplot2)
library(janitor)
library(dplyr)
```


## QUESTION 01: Data Visualisation for Science Communication

### a) Bad Figure Example:

```{r bad figure code, echo = FALSE, warning=FALSE}
# Misleading scatter plot

plot_bad_figure <- ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm, color = species, shape = island)) +
  geom_point(size = 10, alpha = 0.7) +
  labs(title = "Penguin Body Mass vs. Flipper Length",
       x = "body_mass_g",
       y = "flipper_length_mm",
       color = "Species",
       shape = "Island") +
  theme_dark()+
  theme(legend.position = "bottom", legend.title = element_blank(),
        panel.grid = element_blank(), axis.line = element_line(color = "black"),
        plot.title = element_text(hjust = 0.5, size = 20, face = "bold", color = "red"),
        axis.title = element_text(size = 12, face = "italic", color = "purple"),
        legend.text = element_text(size = 8, face = "bold", color = "blue"))

print(plot_bad_figure)

```


### b) Write about how your design choices mislead the reader about the underlying data (200-300 words).

The aims of data visualisation involves four fundamental parts: showing the data; making patterns easy to see; representing magnitudes honestly and drawing graphical elements clearly (Whitlock and Schluter, 2020). The graph above is an example of a bad figure for many reasons. Firstly, the large data point size obscures some of the data. Secondly, the large data point size and colour choice makes it hard to determine the patterns seen in the data.The red and green colour choice is especially bad for those who are colour blind. Additionally, the shape size may influence the reader to misinterpret the magnitude of the data. To elaborate, the size and shape of the data points make it hard to determine where it correlates to on the axis and coupled with the large size of the data points, the reader may infer that the data points have a greater value than they actually do. Finally, the graphical elements are not clear and are distracting to the reader. This includes the numerous colours which are distracting and the dark theme for the background makes it harder to read as do the italicised axis titles. Also, the lack of legend titles makes it harder to interpret the graph. Overall, the graph design is not appealing, hard to read, hides the data and is misleading to the reader, therefore making it difficult to draw valid conclusions.


------------------------------------------------------------------------

## QUESTION 2: Data Pipeline

------------------------------------------------------------------------

### Introduction

This data pipeline aims to load the raw data from the Palmer penguins dataset, clean the said raw data and then perform analyses involving the production of figures. This will involve producing an exploratory figure, running statistical tests and creating a final results figure.


#### Installing / loading packages 

```{r, echo=TRUE, results='hide'}
install.packages(c("ggplot2", "palmerpenguins", "janitor", "dplyr", "ragg","ggpubr"))
library(ggplot2)
library(palmerpenguins)
library(janitor)
library(dplyr)
library(ragg)
library(ggpubr)
```


#### Loading the data 
Creating a new folder called, "data" in my project allows the raw data to be stored in it's unedited form which is beneficial for reproducibility as others can see how the data has been manipulated and it can also be used for others to use in their own analyses. 
```{r}
write.csv(penguins_raw, "penguins_raw.csv")
```


#### Cleaning the data 
I created my own cleaning functions and keep them in another folder called "functions" to prevent repetitive code being written.
```{r, echo=TRUE, results='hide'}

source("functions/cleaning.r") 

#piping multiple functions allows others to easily see the steps 
penguins_clean <- penguins_raw %>% 
  clean() %>%
  shorten_species() %>%
  remove_empty_columns_rows() %>%
  na.omit()

#checking cleaning functions worked 
names(penguins_raw)
names(penguins_clean) 

#save cleaned data 
write.csv(penguins_clean, "data/penguins_clean.csv")
```
#### Creating an exploratory plot 
Now that the data has been cleaned, we can start to visualise the data with exploratory plots.These plots are used to gain a basic understanding of the relationship between variables and therefore do not need to look especially pretty.
```{r Data Exploration, warning=FALSE}

#exploratory plot looking at the relationship between body mass and flipper length
source("functions/plotting.r")

exploratory_plot <- plot_explor(penguins_clean)
exploratory_plot

#saving figure as a vector image so it is scalable 
save_exploratory_plot_svg(penguins_clean,
                          "figures/exploratory_plot.svg",
                          size = 15, scaling = 1)
```


### Hypothesis
The exploratory plot shows that there is a positive correlation between body mass and flipper length. That is to say that as body mass increases, so does flipper length. The plot also shows that species cluster together which may suggesting that there may be a correlation between body mass and flipper length within species.
To look at this, I have separated the species into their own plots so that the effects of species can be seen individually. 

### Statistical Methods
SUbsetting into different species and then performing a linear regression for each will help to identify the statistical significance betweenbody mass and flipper length correlation in the different penguin species. 

```{r Statistics}

#subset into species to look at the linear models for each species 
adelie <- subset(penguins_clean, species == "Adelie")
chinstrap <- subset(penguins_clean, species == "Chinstrap")
gentoo <- subset(penguins_clean, species == "Gentoo")

# fit linear regression models to the different species 
#adelie
adelie_linear <- lm(body_mass_g ~ flipper_length_mm, data = adelie)
summary(adelie_linear) 

#chinstrap
chinstrap_linear <- lm(body_mass_g ~ flipper_length_mm, data = chinstrap)
summary(chinstrap_linear) 

#gentoo
gentoo_linear <- lm(body_mass_g ~ flipper_length_mm, data = gentoo)
summary(gentoo_linear) 

```


### Results & Discussion
The results figure shows the relationship between body mass and flipper length in the three different penguin species. The plot also shows the linear regression for each of the different species along with the Pearson correlation coefficient (R) showing the strength of the correlation and the p-value. 
All three species show significant p-values for these coefficients, suggesting that the linear regression model suggests there is a correlation between body mass and flipper length in the three species.
Additionally, the coefficients vary between species, suggesting there may be species-specific effects between body mass and flipper length correlation. 
To elaborate, in Adelie penguins, the linear regression coefficient (32.689) was shown to be significant (p-value < 3.4e-09). In chinstrap penguins, the linear regression coefficient was 34.573  and also shown to be statistically significant with a p-value of 3.75e-09. Finally, the gentoo penguins showed the strongest correlation between body mass and flipper length with a linear regression coefficient of 54.165 and a p-value of < 2e-16. 

```{r Plotting Results, warning=FALSE}
source("functions/plotting.r")

#explanatory plot 
explanatory_plot <- plot_explan(penguins_clean)
explanatory_plot

#save plot
save_explanatory_plot_svg(penguins_clean,
                          "figures/explanatory_plot.svg",
                          size = 15, scaling = 1)
```

### Conclusion

In conclusion, the results support the hypothesis that there is a positive correlation between body mass and flipper length in  the Palmer Penguins study. Additionally, after noticing clustering of species in the exploratory plot, I subsetted the data to look at whether there were species-specific effects on the correlation between body mass and flipper length and the results also support this, with Gentoo penguins having the strongest correlation of the species. 



------------------------------------------------------------------------







































